Flat-histogram method comparison on 2D Ising model

### 2020-05-01 ising/ising.tex

Liang independently considered whether WL could be treated as a special case of
stochastic approximation whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang et
al.~\cite{liang2007stochastic} argued that WL can be considered a form of
stochastic approximation Monte Carlo (SAMC). Unlike WL, SAMC can guarantee
convergence (if certain conditions are met). Despite the added benefit of
guaranteed convergence, the method still has a system specific user-defined
variable. Such variables often create difficulty when applying Monte Carlo
methods across arbitrary systems.

@@

Pommerenck et al. propose a refinement~\cite{pommerenck2020stochastic} to SAMC
where the update factor is determined dynamically rather than by the user.
stochastic approximation with a dynamic $\gamma$ (SAD) requires the user to
provide the lowest temperature of interest $T_{\min}$. This is analogous to WL
requiring \emph{a priori.} an energy range of interest; however, this is almost always easier to
identify and is more physical than the SAMC parameter $t_0$. The update factor
can be written in terms of the current estimates for the highest $E_H$ and
lowest $E_L$ energies of interest and the last time that an energy in the range
of interest is encountered $t_L$.
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
SAD only explores the energy range of interest as specified by the minimum
temperature of interest $T_{\min} < T < \infty$. During the simulation the two
energies $E_H$ and $E_L$, are refined such that the range of energies are conservatively
estimated. The weights are calculated for each energy region according to the original
prescription.
\begin{enumerate}
\item {$E < E_L$:} $w(E>E_H) = w(E_H)$
\item {$E_L < E < E_H$:} moves are handled the same as other weight-based
methods that are mentioned
\item {$E > E_H$:} $w(E<E_L) = w(E_L)e^{-\frac{E_L-E}{T_{\min}}}$
\end{enumerate}
Each time the simulation changes the value of $E_H$ or $E_L$, the weights
within the new portion of the interesting energy range are updated.

@@

@@

\subsection{The 32 $\times$ 32 Ising model}
Fig.~\ref{fig:N32-gamma} shows the update factor $\gamma_t$ for each of the flat-histogram methods. All of the update factors initially start at $\gamma_{t=0} = 1$. SAD dynamically updates $\gamma_t$ throughout the simulation. After about $10^{10}$ moves, $\gamma^{\text{SAD}}_t$ proceeds as $1/t$. Both WL and ``production'' WL update factors $\gamma^{\text{WL}}_t$ are shown. The production WL begins after $\gamma^{\text{WL}}_t$ has reached $10^{-4}$. The update factor for $1/t$-WL decreases similarly to WL before finding all the energy states and switching to $1/t$. All of the SAMC update factors equal 1 until the number of moves is equal to $t_0$ at which point they decrease as $1/t$.

Fig.~\ref{fig:N32-cv} shows the specific heat capacity vs. the reciprocal temperature $\beta$ at $10^{9}$ moves. Each method (except for the SAMC methods) is shown using the same single random number seed.

Fig.~\ref{fig:N32-cv-error} shows the maximum error in the heat capacity as a function of time for
this system. The solid/dashed lines
represent the average of the maximum value of the error in the specific heat capacity $c_V$ averaged
over eight simulations using different random number seeds. The range of maximum
errors for each simulation is shown as a shaded region. By
the time $10^8$ moves have been made all but the WL simulation have begun to
converge as $1/\sqrt{t}$. We then see the WL error saturate around $10^9$ moves.

>>>>

Liang independently considered whether WL could be treated as a special case of
stochastic approximation whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang \emph{et
al.}~\cite{liang2007stochastic} argued that WL can be considered a form of
stochastic approximation Monte Carlo (SAMC). Unlike WL, SAMC can guarantee
convergence (if certain conditions are met). Despite the added benefit of
guaranteed convergence, the method still has a system specific user-defined
variable. Such variables often create difficulty when applying Monte Carlo
methods across arbitrary systems.

@@

Pommerenck \emph{et al.} propose a refinement~\cite{pommerenck2020stochastic} to SAMC
where the update factor is determined dynamically rather than by the user.
stochastic approximation with a dynamic $\gamma$ (SAD) requires the user to
provide the lowest temperature of interest $T_{\min}$. This is analogous to WL
requiring \emph{a priori.} an energy range of interest; however, this is almost always easier to
identify and is more physical than the SAMC parameter $t_0$. The update factor
can be written in terms of the current estimates for the highest $E_H$ and
lowest $E_L$ energies of interest and the last time that an energy in the range
of interest is encountered $t_L$.
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
SAD only explores the energy range of interest as specified by the minimum
temperature of interest $T_{\min} < T < \infty$. During the simulation the two
energies $E_H$ and $E_L$, are refined such that the range of energies are conservatively
estimated. The weights are calculated for each energy region according to the original
prescription.
\begin{enumerate}
\item {$E < E_L$:} $w(E>E_H) = w(E_H)$
\item {$E_L < E < E_H$:} moves are handled the same as other weight-based
methods that are mentioned
\item {$E > E_H$:} $w(E<E_L) = w(E_L)e^{-\frac{E_L-E}{T_{\min}}}$
\end{enumerate}
Each time the simulation changes the value of $E_H$ or $E_L$, the weights
within the new portion of the interesting energy range are updated.

@@

\subsection{The 32 $\times$ 32 Ising model}
Figure~\ref{fig:N32-gamma} shows the update factor $\gamma_t$ for each of the flat-histogram methods. All of the update factors initially start at $\gamma_{t=0} = 1$. SAD dynamically updates $\gamma_t$ throughout the simulation. After about $10^{10}$ moves, $\gamma^{\text{SAD}}_t$ proceeds as $1/t$. Both WL and ``production'' WL update factors $\gamma^{\text{WL}}_t$ are shown. The production WL begins after $\gamma^{\text{WL}}_t$ has reached $10^{-4}$. The update factor for $1/t$-WL decreases similarly to WL before finding all the energy states and switching to $1/t$. All of the SAMC update factors equal 1 until the number of moves is equal to $t_0$ at which point they decrease as $1/t$.

All the methods except for SAMC gave a reasonable approximation for
the heat capacity peak resulting from the phase transition within 10$^9$ moves.
Figure~\ref{fig:N32-cv} shows the specific heat capacity vs. the reciprocal temperature $\beta$ at $10^{9}$ moves. Each method (except for the SAMC methods) is shown using the same single random number seed.

@@

Figure~\ref{fig:N32-cv-error} shows the maximum error in the heat capacity as a function of time for
this system. The solid/dashed lines
represent the average of the maximum value of the error in the specific heat capacity $c_V$ averaged
over eight simulations using different random number seeds. The range of maximum
errors for each simulation is shown as a shaded region. By
the time $10^8$ moves have been made all but the WL simulation have begun to
converge as $1/\sqrt{t}$. We then see the WL error saturate around $10^9$ moves.
